\chapter{Grammar}

Grammar\index{grammar} is the core of every language. Grammar comprises different themes that are united in a system of changing words enough to build sensible syntax constructions. We will look at Novoslovnica grammar from the point of the parts of speech.

\newglossaryentry{pos}{name=POS, description={Part of speech}}

\textit{Part of speech\index{part of speech} (\gls{pos})} - is a category of words which have similar grammatical properties.

Thus, we unite a group of words into parts of speech when they have a lot of similar grammar properties such as “number”, “person”, “case”, “tense” etc. In this chapter we will go through different parts of speech and study what differences they have and how we should identify each of them and combine them to be able of creating phrases and sentences. 

Parts of speech comprise two main categories - independent\index{independent POS} POS and auxiliary\index{auxiliary POS} POS. The fact the POS is independent or not refers to its semantic value. An independent POS has a full semantic value that can be used separately.  That means when you say a word of an independent POS you reproduce some semantic meaning that the interlocutor can recognize. An auxiliary POS a semantic value that is partially defined and can be distinguished only in pair with a word of an independent POS.

\textbf{Example:}

\textit{River} - this word can be recognized by interlocutor as some concept of water flowing in a restricted area.

\textit{Beautiful} - this word is recognized as an attribute of any concept that is nice, pleasant to the person (i.e. speaker or interlocutor)

\textit{Came} - this word can be recognized as any concept’s action of going to the destination point and having reached it. 

And so on. Nouns, adjectives, verbs, adverbs, participles, numerals, pronouns refer to independent POS. You see that every word has its own meaning full enough to imagine yourself some information you have received. This is not true for a word of an auxiliary POS. 

\textit{On} - this word can be recognized as a placement of an object against some horizontal surface. It can refer to some time moment (be on time). It can also be used with a fully different value (come on - to increase the activity of doing something).

\textit{For} - it can be recognized as an aim for something or an object that participates in some action. Also this word can refer to a duration of a process.

Moreover, words “and” or “to” cannot give you any map in your mind into any sensible meanings. Though these words have no determined semantic value, they are extremely important in the whole phrase connecting two words of the same POS with logical value (AND, OR, NOT). English language is an analytical one, that is why words are mostly connected with each other in the phrase by an auxiliary POS. Without them you are not able to understand what the person is speaking about. Slavic languages are fusional, however, there are enough analytic features in them, hence auxiliary parts of speech are also important. Articles, prepositions and conjunctions are referred to auxiliary parts of speech.

There are also two additional categories - particles and interjections. Some allocate them into separate categories, some claim they belong to an auxiliary category. Nevertheless, these are both separated parts of speech because they have different grammar properties.

These are all categories of POS. If we speak about an independent POS, we should take into account that there are different semantic, morphological and syntax functions can be described by it. There are several types of semantic functions: the concept, the attribute, the predicate and the demonstrative.

The \textit{concept} is something that correlates with the object or subject in the real world. It could be either abstract or imaginary, but we can ask a questions “Who? What?” about it.

The \textit{predicate} is something that determines the action, corresponding to a concept. We often ask questions “What to do?” to reveal a predicate.

The \textit{attribute} is something that is correspond to a concept or an action. We ask questions “What concept is like?” or “What action is like?” to find out the determine value of the attribute.

The \textit{demonstrative} points out the concept. It has the same question with the attribute yet has no semantic value but demonstrating the concept it corresponds with. 

Parts of speech that have properties of a concept are: \textit{nouns, cardinal numerals, verbal noun and some kinds of pronouns}.

Parts of speech that have properties of an attribute are: \textit{adjectives, participles, ordinal numerals and adverbs}.

Parts of speech that have properties of a predicate are: \textit{verbs, transgressive, and gerund}.

Parts of speech that have demonstrative properties are most kinds of \textit{pronouns}.

Thus, noun, adjective, verb, adverb, numeral, pronoun, gerund and participle are independent POSes, while preposition, particle, interjection (with Onomatopoetic), article and conjunction are auxiliary ones.

Independent parts of speech are also divided into nominal and verbal ones. It is extremely important because this division shows differences in grammar forms of nominal and verbal POS. Verb, participle, transgressive, gerund are verbal parts of speech, while noun, adjective, numeral are nominal. Adverb and pronoun stay separately - the first one because of its immutability and the second one because of its heterogenity. 

In the chapter it will be spoken about the very POS exists in Novoslovnica. The table with grammar and semantic properties of a POS will be given in the corresponding section.

First of all you should know some facts about different grammatical properties in Novoslovnica.

\input{./content/case3.1}
\input{./content/number3.2}
\input{./content/person3.3}
\input{./content/tense3.4}
\input{./content/mood3.5}
\input{./content/noun3.6}
\input{./content/adjective3.7}
\input{./content/pronoun3.8}
\input{./content/numeral}
\input{./content/adverb3.9}
\input{./content/predicative3.10}
\input{./content/verb3.11}
\input{./content/participle3.12}
\input{./content/gerund3.13}
\input{./content/article3.16}
\input{./content/particle3.17}
\input{./content/preposition3.18}
\input{./content/conjunction3.19}
\input{./content/interjection3.20}
\input{./content/onomatopoetic3.21}


