\section{Gerund}

Gerund\index{gerund} is a form of a verb that determines the process of an action. That means it is not an independent POS, but has some features that do not allow us to include the description about gerund in the paragraph about verb.

Gerund is formed from the infinitive of a verb by reducing a “-ti” ending and adding a suffix “-n-” and an ending “-e-”. Gerunds are often created only from verbs of A-type. However, you can construct gerunds from verbs of other types. Look at the examples:

\textbf{Examples:}

\textit{Pisati} (verb) - pisane (gerund) - To write/Writing

\textit{Nositi} (verb) - nosine (gerund, bad form) - nošane (gerund, recommended form) - To wear/Wearing

\textit{Pěti} (verb) - Pěne (gerund, bad form) - pějane (gerund, recommended form) - To sing/singing

Unlike infinitive or supine, gerund can be declined, but only in the singular. This fact unites it with the nominal POS. Simply speaking, gerund declines so as soft neutral nouns do despite the nominal/accusative form of “e”.

\begin{table}
	\begin{tabular}{ll}
	Case & Ending \\
	Nominal & -e \\
	Genitive & -ä \\
	Partitive & -ü \\
	Dative & -ü \\
	Accusative & -e \\
	Instrumental & -ëm \\
	Prepositional & -ě \\
	Locative & -ü \\
	Vocative & no form
	\end{tabular}
\end{table}

Let us look at some examples:

\textbf{Examples:}

\textit{Ja dobro znam pisane} - (Accusative) I am good at writing

\textit{Vlådam dobro pisanëm} - (Instumentative) My writing is good.

\textit{Ja ne pravim grěšek v pisanü} - (Locative) I do not mistake while writing

\subsection{Gerund vs Verbal noun vs Infinitive}

All of these three POSes describe a process. However, they are different (Gerund, verb form and noun form) so it is necessary to define the properties of each other.

Verbal noun describes a process as an object. Often it corresponds with the result of the action (\textit{pisanije} - the result of writing). It has a role of Subject or Object.

Gerund describes a process as an action (\textit{pisane} - the process of writing something). It also has a role of Subject or Object.

Infinitive represents a process itself as a definitive normal form (pisati - to write). It is used mostly in complex verbal forms.

\textbf{Examples:}

\textit{Ove pisanije je mnogo lěpo} - It is a beautiful writing.

\textit{Pisane je mnogo trudno dlä menę} - Writing is very hard for me.

\textit{Hoču pisati knigu} - I want to write a book.

\subsection{Gerund vs Transgressive}

Both Transgressive and Gerund are described through an "-ing" form in English. Therefore, it should be useful to have some distinguising rules to classify the verb forms in Novoslovnica.

Transgressive and gerund have different syntax roles (look for them in the next chapter). The following table shows which roles they can have.

\begin{table}
	\begin{tabular}{lll}
		Role & Transgressive & Gerund \\
		Subject & - & + \\
		Object & - & + \\
		Modifier & + & - \\
	\end{tabular}
\end{table}

Thus, Transgressive has an attributive value describing a predicate with an additional action as its modifier. Gerund represents the process itself so it has a subjective value while describing a process.