\section{Noun}

\begin{table}[h]
	\caption{Noun characteristics}
	\begin{tabular}{lp{12em}}
		\textbf{Title}              & \textbf{Value}                            \\
		Semantic value              & Concept                                   \\
		Category                    & Independent                               \\
		Subcategory                 & Nominal                                   \\
		Alteration                  & Declension                                \\
		Alteration parameters       & Case, Numbers, Gender, Type of Declension \\
		Differentiation parameters  & Gender, Animacy, Types of  Declension                                  
	\end{tabular}
\end{table}

\newglossaryentry{noun}{name=N, description={Noun}}


Nouns\index{noun} (\gls{noun}) can be differentiated by three parameters: gender, animacy and the type of declension.

\underline{Animacy\index{animacy}} determines whether the object is animate (we are able to ask “Who is it?” to the object) and inanimate (we are able to ask “What is it?” to the object).

\underline{Gender\index{gender}} determines whether the object is masculine, feminine or undefined (we cannot say it is one of the previous genders). Hence, there are three genders: masculine (with masculine properties), feminine (with feminine properties) and neutral (with undefined properties).

Despite English, Novoslovnica make us always show word gender explicitly of both animacies. We can say “it” to the object if we aren’t coupled with it in English. In Novoslovnica (as in every Slavic language) we should use the predefined gender when we speak about some concept (noun). Using wrong genders shows your ignorance and language nescience.

\underline{Type of declension} is a parameter of declension function. Declension\index{declension} is a function of word alteration. It has two input parameters - the word itself and the type of declension that includes the terms of animacy, gender and some morphological features (such as word endings) in it. The output is a list of forms that the noun can be changed into. Novoslovnica supports 27-cell output list with (3 numbers) * (9 cases) elements in it.

Further you can see tables of different declension types. The tables cover all use cases of declension function. You can find all of them in the eighth chapter.

\newglossaryentry{animate}{name=An, description={Animate}}
\newglossaryentry{inanimate}{name=Inan, description={Inanimate}}

P.S. In all tables abbreviations “\gls{animate}” and “\gls{inanimate}” are for “animate” and “inanimate” respectively.


\subsection{Collective nouns}

Collective nouns denote a set of homogeneous objects or living beings as an indivisible whole.

Collective nouns have special grammatical features:

\begin{itemize}
	\item Number immutability, only a singular form exists.
	\item Incongruity with quantitative numerals and with words denoting units of measure.
	\item Can be combined with the words \textit{many/few} or \textit{how many} in the singular form.
\end{itemize}

Collective nouns are indicated with the following suffixes:

\begin{itemize}
	\item ij
	\item stv
\end{itemize}

\textbf{Examples:}

- \textit{Bratija} - Brethren

- \textit{Pijonerija} - Pioneers

- \textit{Bratstvo} - Brotherhood

- \textit{Čelověkstvo} - Mankind


\input{./content/verbal-noun.3.15}
