\section{Pronoun}

\begin{table}[!hbt]
	\caption{Pronoun characteristics}
	\begin{tabular}{lp{12em}}
		\textbf{Title}              & \textbf{Value}               \\
		Semantic value              & Attribute, Concept           \\
		Category                    & Independent                  \\
		Subcategory                 & Nominal                      \\
		Alteration                  & Declension                   \\
		Alteration parameters       & Case, Numbers, Gender, Person\\
		Differentiation parameters  & Gender, Type, Group
	\end{tabular}
\end{table}

Pronoun\index{pronoun} is a POS that has different meanings. It can play the role of an attribute or a concept, depending on what is replaced with the pronoun. However, pronoun has a verbal property of person.

Pronouns are divided into three types: nominal\index{pronoun!nominal} (noun-like declension), substantive\index{pronoun!substantive} (substantive declension), attributive\index{pronoun!attributive} (adjective-like declension).

There are also several groups of pronouns, depending on their semantic value. They are: personal, possessive, interrogative, relative, indefinite, definitive, reflexive, demonstrative, negative, reciprocal.

\subsection{Personal pronouns}

Personal pronouns\index{pronoun!personal} are one of the most important groups in the Pronoun POS. They replace nouns in sentences when we try to avoid an unreasonable repeat. Personal pronouns have different forms for every person-number cell. In the table you can see them.

\begin{table}[!htb]
	\begin{tabular}{llll}
		& Singular & Dual & Plural \\
		1 person & Ja (Az) & Ma & My \\
		2 person & Ty & Va & Vy \\
		3 person, ms & On & Ona & Oni \\
		3 person, fm & Ona & One & Oně \\
		3 person, neu & Ono & Oná & Onji
	\end{tabular}
\end{table}

I should mention the form “Vy” (You). As in English or polite Russian, we can use this pronoun for a single person when we want to emphasize our respect to the interlocutor. Moreover, we also have to use a plural form of the verb while speaking “Vy” in polite form.

Also you can see two forms of the 1 person - singular pronoun (I). Etymologically the form “Az” is full while “Ja” is only a short one. However, different Slavic languages have retained different forms and now we see Novoslovnica should support both ones. In Novoslovnica the difference between using “Az” and “Ja” lies in phonetics. As you see, “Az” begins with the vowel and ends with a consonant, “Ja” - controversially. If we remember rule 1 we will get such a list of rules:

\begin{itemize}
	\item If the word before the pronoun ends with a vowel and the word after begins with a consonant - you should use “Ja”
	\item If the word before the pronoun ends with a consonant and the word after begins with a vowel - you should use “Az”
	\item In other cases you can use either one or another variant\footnote{“Az” is more formal than “Ja”. In everyday speech you should use "Ja" variant while in scientific, official only "Az" should be used.}
\end{itemize}

Now let us speak about declension of personal pronouns. Personal pronouns relate to nominal pronouns (with noun-like declension). This is the most difficult type to learn.\footnote{The declension of personal pronouns is derived mostly from old-Slavonic language except for 1 person dual. The similar declension within still languages can be found in Slovenian.} But everything has its order and beauty.



\subsection{Reflexive pronoun}

This\index{pronoun!reflexive} is a separate group of pronouns. There is the only reflexive pronoun in Novoslovnica - “sebę”. Its feature is the absence of nominative form. It has only 7 cases to be alternated. Vocative and Nominative are absent.

\begin{table}[!htb]
	\begin{tabular}{lll}
		Case & Full & Short \\
		Genitive & Sebę & Sę \\
		Partitive & Sebä & Sä \\
		Accusative & Sebe & Se \\
		Dative & Sebi & Si \\ 
		Insrumental & Sobom (Soboǐ) & - \\ 
		Prepositional & O sobě & - \\
		Locative & V sobu & -
	\end{tabular}
\end{table}


“Sę” is used very often as a reflexive suffix in verbs. It determines a way of creating medial voice sentences (look the paragraph about it).

However, there is a term of a complex reflexive form (CRF) also. It is formed by the sequence of a personal pronoun and the definitive pronoun “sám”. This form shows a partly-developed reflection of a subject.  

\textbf{Examples:}

\textit{On sám stroji búdu-ta.} -

\textit{On sę stroji.}

\textit{Búda-ta sę stroji nim.}

There is practically no difference in using a reflexive pronoun or a complex reflexive form. However, it is recommended to us a CRF for Nominative and to use a reflexive pronoun in other cases.

\subsection{Possessive pronouns}

Possessive\index{pronoun!possessive} pronouns show whom any object belongs to. For example, we can say “\textit{It is a thing of Bob} (which Bob possesses)”. In Novoslovnica you can use a similar expression: “\textit{To je věčj Boba}” (remember rules of case using). Also Novoslovnica has another expression with a possessive adjective: “To je Bobóva věčj”. This adjective shows that this thing belongs to Bob. However, if we have already used the name of Bob in our sentence, we should not repeat it again. English also uses possessive pronouns in such cases: “\textit{Bob is my friend and that’s his thing}”. We do not repeat the word Bob, we just say - his (which he possesses). That is what the possessive pronouns look like.

These pronouns are attributive, so we have no need to rewrite their declension, just to name nominative forms. Further, you take a nominative form of a possessive pronoun, look through declension tables for adjectives and transform your pronoun so as you did it with an adjective. Now let us look at the nominative forms of possessive pronouns.

\begin{table}[!htb]
	\caption{Possessive pronouns}
	\begin{tabular}{llll}
		& Singular & Dual & Plural \\
		1 person & Môǐ & Naš & Naš \\
		2 person & Tvôǐ & Vaš & Vaš \\
		3 person, ms & Jegôǐ & Onyš & Ih \\
		3 person, fm & Jejôǐ & Oneš & Ih \\
		3 person, neu & Jegôǐ & Onyš & Ih
	\end{tabular}
\end{table}

There also exists possessive adjectives you can see on the next table.

\begin{table}[!htb]
	\caption{Possessive adjective}
	\begin{tabular}{llll}
		& Singular & Dual & Plural \\
		1 person & Môǐnyǐ & Našnyǐ & Našnyǐ \\
		2 person & Tvôǐnyǐ & Vašnyǐ & Vašnyǐ \\
		3 person, ms & Jegôǐnyǐ & Onyšnyǐ & Ihnyǐ \\
		3 person, fm & Jejôǐnyǐ & Onešnyǐ & Ěhnyǐ \\
		3 person, neu & Jegôǐnyǐ & Onyšnyǐ & Ihnyǐ
	\end{tabular}
\end{table}

Possessive adjectives are used only in full form to underline the importance of something. Usually, possessive pronouns should be used.

Look at the examples to see the differences.

\textit{Môǐ pes lübi męso} - My dog likes meet.

\textit{Môǐnyǐ pes lübi męso} - My dog (that we are talking about now) likes meet.

Possessive forms (adjectives and pronouns) should be placed right before the identified word. Nevertheless, you can use a postfix form to describe possessiveness. In that case, you should use personal pronoun in Dative in role of a possessive one. Look at the examples.

\textit{Daǐte moju knigu - Daǐte knigu mi} (the latter is not recommended due to possible misunderstanding)

\textit{Ja  sòm poslal syna mi do pěkarnï - Ja sòm poslal mojoga syna do pěkarnï.}

As you see, such a form does not decline along with the main word. In the following table you can find the correlations between declinable prefix form and a "frozen" postfix one.

\begin{table}[!htb]
	\begin{tabular}{ll}
		\textbf{Prefix} & \textbf{Postfix} \\
		Môǐ & Mi \\
		Tvôǐ & Ti \\
		Svôǐ & Si \\
		Jegôǐ & Mu \\
		Jejôǐ & Ji \\
		Naš & Nam / Nama \\
		Vaš & Vam / Vama \\
		Jih & Jim \\
		Onyš & Onama \\
		Oneš & Oněma \\
	\end{tabular}
\end{table}

Note that the postfix form is pretty similar with use of genitive. The dative possessive form is used mostly in fiction texts to underline the beauty of writing. However, dative for singular pronouns is more frequently used than the genitive one.

\textbf{Examples:}

\textit{To je moja kniga} (Possessive pronoun) - Frequently used

\textit{To je kniga mi} (Dative) - Frequently used

\textit{To je kniga mę} (Genitive) - Rarely used

\textit{On je syn naju} (Genitive) - Frequently used

\textit{On je naš syn} (Possessive pronoun) - Frequently used

\textit{On je syn nama} (Dative) - Rarely used

\subsection{Interrogative and relative pronouns}

Interrogative\index{pronoun!interrogative} pronouns are used when we create a question. They are never in plural or dual. Also they have no declension. The only role of interrogative pronouns is to reveal an aim of the question (How much? Who? What?) - to guide the interlocutor to the right answer (you need).

Relative\index{pronoun!relative} pronouns aim is to introduce a relative clause.

\textit{Relative clause}\index{clause!relative} is a special kind of subordinate clause whose primary function is as modifier to a noun or nominal. We examine the case of relative clause modifiers in NPs first, and then extend the description to cover less prototypical relative constructions.\cite{english-grammar}

Interrogative and relative pronouns are practically the same, only usage differs, that is why I talk about them in the same paragraph. Let us look at them in the table.

\begin{table}[!htb]
	\begin{tabular}{lll}
		English equivalent & Interrogative pronoun & Relative pronoun \\
		Who & Kto? & Ke \\
		What & Čto & Če \\
		What & Kakyǐ? & Kak \\
		Which & Ktoryǐ? & Ktor \\
		Whose & Čyǐ? & Čyǐ \\
		What & Jakyǐ? & Jak \\
		What & Kakvyǐ ? & Kakòv \\
		What & Kolïkyǐ ? & Kolïk 
	\end{tabular}
\end{table}

Pronouns “Kto”, “Čto” have a similar declension with possessive pronouns. Other pronouns decline so as adjectives do.
  
\subsection{Indefinite and negative pronouns}

Another group of pronouns that are similar to each other are indefinite\index{pronoun!indefinite} and negative\index{pronoun!negative} pronouns. They are derived from interrogative pronouns by adding the negative “ni-” and the indefinite “ne-” prefixes. Their declension equals to their ancestor’s one.

\begin{table}[!htb]
	\begin{tabular}{lll}
		Interrogative pronoun & Negative pronoun & Indefinite pronoun \\
		Kto? & Nikto & Někto \\
		Čto & Ničto & Něčto \\
		Kakyǐ? & Nikakyǐ & Někakyǐ \\
		Ktoryǐ? & Niktoryǐ & Něktoryǐ \\
		Čyǐ? & Ničyǐ & Něčyǐ 
	\end{tabular}
\end{table}

\subsection{Demonstrative pronouns}

Demonstrative\index{pronoun!demonstrative} pronouns are usually used to make the interlocutor pay attention to something.
There are only four demonstrative pronouns. And this topic is closely connected with the articles. In the table 4.20 you can see these pronouns, divided by the terms of visibility and distance of the object with respect of the speaker.


\begin{table}[!htb]
	\caption{Demonstrative pronouns}
	\begin{tabular}{lll}
		& Visible & Invisible \\
		Far & Onyǐ & Tyǐ \\
		Close & Ovyǐ & Sïǐ  
	\end{tabular}
\end{table}

These pronouns decline so as adjectives do, so that is very easy. Nouns that are used with demonstrative pronouns should not be determinized with article. You should use \underline{either} demonstrative pronoun or articles.

\textbf{Examples:}

\textit{Onyǐ stol je starym} - That table (you are looking at) is old.

\textit{Stol-òn je star} - The table is old.

\textit{Ova kniga je korystnoǐ} - This book is interesting.

\textit{Kniga-va je korystna} - The book is interesting.

\subsection{Attributive pronouns}

Attributive\index{pronoun!definitive} pronouns indicate a generalized feature of an object. 

\begin{table}[!htb]
	\caption{Attributive pronouns}
	\begin{tabular}{ll}
		English equivalent & Definitive pronoun \\
		Whole & Vesj \\
		All & Vsë, vsä, vsï \\
		Every & Vsękyǐ, lübyǐ \\
		Each & Káždyǐ \\
		Another, other & Ïnyǐ, drugyǐ \\
		Both & Oba, obě \\
		Reflexive pronouns & Personal pronouns + sám
	\end{tabular}
\end{table}

All these pronouns decline as adjectives. I should say just about the last one - the pronoun “sám”. This pronoun is used with personal pronouns to create a complex reflexive form (look paragraph about reflexive pronouns). However, it keeps adjective-like declension.

\subsection{Reciprocal pronouns}


The last type of pronouns is reciprocal\index{pronoun!reciprocal}. It is used to refer to a noun phrase mentioned earlier in a sentence. English has only two such pronouns - each other and another one.

Slavic languages have much more variants:

\textit{Drug so drugom} - With each other

\textit{Drug drugu} - Each other

\textit{Raz za razom} - Time by time

