\section{Stem}

A word \textit{stem} is a semantic core of the word. It includes the main semantic value presented by a word form.

Each word of an independent \gls{pos} has at least one stem.

A \textit{stem nest} is a group of words existing in the language that have the same word stem. For example, look through the next words:

\textit{Sųd - sųdba - sųditi - sųden}

\textit{Hod - hodka - hoditi - poholdka - zahoden}

You see the words are of different meanings and different \gls{pos}. Thought, every word in each group has an equal semantic core - something connected with the judgement (1) or with walking (2).

Stems can be divided by several categories. For examples, there are borrowed and native stems. Native stems are derived from some proto-language forms. Thus, words which stems are derived from proto-Slavic, proto-Baltic, proto-German are supposed to be native in Novoslovnica. Also, some neologisms created by Slavic languages are also native.

Words of all other stems are borrowed. On of Novoslovnica's goals is to reduce the borrowings. However, we can list the languages from which Novoslovnica borrows the most. We use the term "factor" to describe the affect on Novoslovnica lexicon of different non-Slavic languages instead of absolute or percentage value. The more the factor the more influence on the language the borrowing from that language have.

\begin{table}[!htb]
	\begin{tabular}{ll}
		Language & Factor \\
		Roman & 43 \\
		i.e. French & 2 \\
		i.e. Latin & 40 \\
		German & 26 \\
		i.e. English & 25 \\
		Greek & 12 \\ 
		Turkish & 6 \\
		Baltic & 0.01 \\
	\end{tabular}
\end{table}

The word can have multiple stems. Novoslovnica usually use words with up to three stems in a word. The stems are connected with stem-forming vowels. They are: \textit{o, ë, ï} and \textit{ô}.

Words with multiple stems usually describes a \gls{np} that has a strong forces between the words. The following examples show when you should use each of stem-forming vowels.

\textbf{Examples:}

- \textit{Zemlëtręs} - \textit{Tręs zemlï} - Earthquake (auxiliary word (soft) + main word)

- \textit{Sámovoz} - \textit{Vozi sámo} - Automobile (auxiliary word (hard) + main word)

- \textit{Blågodariti - Blågo dariti} - To thank (auxiliary word (hard) + main word)

- \textit{Pętïkrátno} - \textit{Pętï krátno} - Five times (Simple concatination of numerals.\footnote{Pętokrátno is also valid according to the second case (Kpátno pętï - Pętokrátno)})

- \textit{Dvôkrátno - Dva kráta} - Twice (Particular case of the previous example.\footnote{Dvokrátno from Krátno dvôm is also available.})
