\subsection{Transgressive}

Transgressive\index{transgressive} is a form of a verb so as infinitive and supine are.

Transgressive cannot be declined, it has verbal characteristics (i.e. forms) but it is derived from participle with its suffixes. Let us look at the table with transgressive forms.

\begin{table}
	\begin{tabular}{lll}
		& Imperfect form & Perfect form \\
		Suffix & -ja (-čï) || -jučï & -v (-šy)
	\end{tabular}
\end{table}

We see that there are no endings here, that means transgressive cannot be declined. Perfect form determines the action that is completed (in past, present or future). Imperfect form determines the action that is being done (in past, present or future). This fact differ transgressives from participles, because, as you can see, participles are divided by tenses, not by forms. However, transgressive is “over” the tenses. 

It should be noted that a consonant “j” in imperfect form transforms into a soft symbol after consonants. 

\textbf{Examples:}

\textit{Spaja - Spä} (Sleeping)

\textit{Vëdaja - Vëdä} (Leading)

There are recommendations about using different forms of transgressive. Perfect forms with “-v-” are used when we speak about actions in the past (Aorist), while forms with “-všy-” are used when we speak about resultative action by the present moment (Perfect). Imperfect full forms with “-jačï-” are less preferred than with “-jučï-”. Moreover, these full forms are recommended to be used as a reflection of actions in the past (Imperfect), while short form of “-ja-” should be used as a reflection of actions in Present Concrete tense.

You can see the recommendations in the table.

\begin{table}
	\begin{tabular}{lll}
		Tense & Suffix & When/while mark \\
		Present Definite & -ja & while \\
		Imperfect & -jučï & while \\
		Aorist & -v & when \\
		Perfect & -všy & when
	\end{tabular}
\end{table}

Transgressive can be replaced with an additional relative clause in the sentence of the tense and when/while mark corresponding with the previous table.


Look at the \textbf{examples}:

\textit{Kazavšy mu něčto, toǐ izidaše iz doma.} - Said something to him, he came out of the house.

\textit{Koĝda toǐ kazal byše mu něčto, on izidaše iz doma.} - When he had said something to him, he came out of the house.

