\section{Verb}

Verb\index{verb} is a POS that describes the predicate and its properties. 


\subsection{Verbal types}

Learning Slavic languages you could mention that there is a set of verb suffixes that is very eliminated. Novoslovnica provides a theory that allows you to construct a right verb form.
Verbs with different verb suffixes represent different verbal types. There are four types of verbs in Novoslovnica.

\begin{itemize}
	\item A-type\index{verb!a-type}: verbs of this type define that the action in common sense.  
	\item E-type\index{verb!e-type}: verbs of this type define that the action is long-termed.
	\item I-type\index{verb!i-type}: verbs of this type define that the action is short-termed. This type comprises verbs with suffixes I and O. The suffix O is used when the consonant before the constructed suffix is involved in alterations depending on soft vowels that the vowel I is.
	\item U-type\index{verb!u-type}: verbs of this type define that the action is dotty.
	\item Extra type\index{verb!extra-type}. Is formed with the suffix “-OVA-” and defines the repeated action.  
\end{itemize}

When you speak about the action, you find what characteristic is suitable for the action and then use one of the predefined verbal types.

Some can ask what are the differences in tenses and types, because it might confuse you. However, verbal type determines the durability of the action (or its repeating property) while tense determine tense characteristic of the action such as completeness, stability in time, result, order etc.

All tenses provides difference between conjugation of different verbal types except imperfect. This tense have the common conjugation table for all verbal types. 

Further we will speak about conjugation itself. Be ready for a lot of tables! We will look at verb conjugation in indicative mood first of all. Then we will speak about other moods.

\subsection{Active voice}

Active\index{voice!active} voice shows that the person makes the action by himself. So, the subject of the sentence and the actor are the same.

\subsubsection{Indicative mood}

Verbs in indicative\index{mood!indicative} mood can be found in every tense that novoslovnica possesses. Let us look at some tables with verb of different verbal types conjugation.

\textbf{Present Tenses}

\textbf{A-type conjugation}

\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Common & Singular & Dual & Plural \\
		1 person & -am & -ama & -ame \\
		2 person & -aš & -ata & -ate \\
		3 person & -a & -at & -ut
	\end{tabular}
\end{table}


\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Concrete & Singular & Dual & Plural \\
		1 person & -aju & -ajema & -ajeme \\
		2 person & -aješ & -ajeta & -ajete \\
		3 person & -aje & -ajat & -ajut
	\end{tabular}
\end{table}

\textbf{E-type conjugation}

\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Common & Singular & Dual & Plural \\
		1 person & -em & -ema & -eme \\
		2 person & -eš & -eta & -ete \\
		3 person & -e & -at & -ut
	\end{tabular}
\end{table}


\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Concrete & Singular & Dual & Plural \\
		1 person & -u & -ujema & -ujeme \\
		2 person & -uješ & -ujeta & -ujete \\
		3 person & -uje & -ujat & -ujut
	\end{tabular}
\end{table}

\textbf{I-type conjugation}

\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Common & Singular & Dual & Plural \\
		1 person & -im & -ima & -ime \\
		2 person & -iš & -ita & -ite \\
		3 person & -i & -at & -ut
	\end{tabular}
\end{table}


\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Concrete & Singular & Dual & Plural \\
		1 person & -ujim & -ujima & -ujime \\
		2 person & -ujiš & -ujita & -ujite \\
		3 person & -uji & -ujat & -ujut
	\end{tabular}
\end{table}

\textbf{U-type conjugation}

\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Common & Singular & Dual & Plural \\
		1 person & -nam & -nama & -name \\
		2 person & -naš & -nata & -nate \\
		3 person & -na & -nat & -nut
	\end{tabular}
\end{table}


\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Concrete & Singular & Dual & Plural \\
		1 person & - & - & - \\
		2 person & - & - & - \\
		3 person & - & - & -
	\end{tabular}
\end{table}

\textbf{Extra-type conjugation}

\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Common & Singular & Dual & Plural \\
		1 person & -ovam & -ovama & -ovame \\
		2 person & -ovaš & -ovata & -ovate \\
		3 person & -ova & -ovat & -ovut
	\end{tabular}
\end{table}


\begin{table}[!htb]
	\begin{tabular}{llll}
		Present Concrete & Singular & Dual & Plural \\
		1 person & -uju & -ujema & -ujeme \\
		2 person & -uješ & -ujeta & -ujete \\
		3 person & -uje & -ujat & -ujut
	\end{tabular}
\end{table}


\textbf{Exceptions:}

\begin{table}[!htb]
	\caption{The verb "Byti"}
	\begin{tabular}{lllllll}
		Pr. Common
			& \multicolumn{2}{c}{Singular}
			 & \multicolumn{2}{c}{Dual}
			 & \multicolumn{2}{c}{Plural} \\
		1 person & Jesòm & Sòm & Jesma & Sma & Jesme & Sme \\
		2 person & Jesi & Si & Jesta & Sta & Jeste & Ste \\
		3 person & Jestj & Je & Jesų & Sų & Jesu & Su
	\end{tabular}
\end{table}


\textbf{Future Tenses}

\textbf{Future Definite Tense (Bųdešt čas)}

\textbf{Future II}

\textbf{Future-in-the-Past I}

\textbf{Future-in-the-Past II}

\textbf{Past Tenses}

\subsubsection{Subjunctive mood}

Subjunctive\index{mood!subjunctive} mood shows two states of an actions. In the first state it can show a desirable action. In the second one it can show an action that has not become a real one. 

Subjunctive mood has only one-tense form. It is constructed with the verb “byti” in a subjunctive form with an aorist or imperfect participle. Special forms of the verb “byti” you can see in the next table.

\begin{table}[!htb]
	\begin{tabular}{llll}
		Subjunctive mood & Singular & Dual & Plural \\
		1 person & Bih & Bihma & Bihme \\
		2 person & Biša & Bihta & Bihte \\
		3 person & Biše & Biha & Bihu
	\end{tabular}
\end{table}

\subsubsection{Imperative mood}

Imperative\index{mood!imperative} mood shows that the actor make somebody do some action (imperate another object). Imperative mood also has only one tense to be used in. 

\begin{table}[!htb]
	\caption{A-type}
	\begin{tabular}{llll}
		Imperative mood & Singular & Dual & Plural \\
		1 person &  & -aǐma & -aǐmo \\
		2 person & -aǐ & -aǐta & -aǐte \\
		3 person &  &  & 
	\end{tabular}
\end{table}



\begin{table}[!htb]
	\caption{E-type}
	\begin{tabular}{llll}
		Imperative mood & Singular & Dual & Plural \\
		1 person &  & -eǐma & -eǐmo \\
		2 person & -i & -eǐta & -eǐte \\
		3 person &  &  & 
	\end{tabular}
\end{table}



\begin{table}[!htb]
	\caption{I-Type}
	\begin{tabular}{llll}
		Imperative mood & Singular & Dual & Plural \\
		1 person &  & -iǐma & -iǐmo \\
		2 person & -i & -iǐta & -iǐte \\
		3 person &  &  & 
	\end{tabular}
\end{table}

\begin{table}[!htb]
	\caption{U-Type}
	\begin{tabular}{llll}
		Imperative mood & Singular & Dual & Plural \\
		1 person &  & -niǐma & -niǐmo \\
		2 person & -ni & -niǐta & -niǐte \\
		3 person &  &  & 
	\end{tabular}
\end{table}


\begin{table}[!htb]
	\caption{Extra-Type}
	\begin{tabular}{llll}
		Imperative mood & Singular & Dual & Plural \\
		1 person &  & -uǐǐma & -uǐǐmo \\
		2 person & -uǐ & -uǐǐta & -uǐte \\
		3 person &  &  & 
	\end{tabular}
\end{table}

\subsubsection{Inferential mood}

Inferential\index{mood!inferential} mood is used when we speak about actions that we did not observe by ourselves. This mood mostly is used in tales, histories two show that we were not witnesses of what we are speaking about.

Inferential mood is created by using the verb “ïmáti’ with the past passive participle (PPP). This mood can be used in all tenses of the indicative mood (we should only place the verb “ïmáti” into this tense and add a PPP of the main verb to it). 

\textbf{Examples:}

\textit{Ja sòm rođen svojoǐ mamoǐ} - I am born by my own mother. (This is a real fact, but the actor (Mother) is not at the subject place in the sentence - Passive Voice).

\textit{Ona má rođeno ove rebętko.} - Some say she bore this child (This is a rumor, and the actor is in the place of a subject - Inferential mood of AV)

\subsection{Passive voice}

Passive\index{voice!passive} voice shows that the person is an object for the action. That means, that the subject of the sentence semantically is not an actor, but the object. 

Passive voice has only two moods - Indicative and Subjunctive. Passive voice is created by using the verb “byti” with the PPP. Here is the narrow border between Inferential mood of Active Voice and Indicative mood of Passive Voice.

We can describe the difference between these two terms in such a way. If we speak about real actions that we have imagined or we heard about them (however, the actor of these sentences is placed in the subject) - we use Inferential mood. And if we speak about any action that has no actor in the subject syntax role - we use Passive Voice. Let’s look at the examples.


\input{./content/infinitive_supine}


